Language learning via recursive pragmatic reasoning

### 2013-05-31 praglearn.tex
cnclusion

Two puzzles present themselves to language users: What do words mean in general and what do they mean in context?
%Natural language is a powerful tool for conveying meaning in context. 
Consider the utterances
``it's raining,'' ``I ate some of the cookies,'' or ``can you close
the window?'' In each, a listener must go beyond the literal meaning
of the words to fill in contextual details (``it's raining here and
now''), infer that a stronger alternative is not true (``I ate some
but not all of the cookies''), or more generally infer the speaker's
communicative goal (``I want you to close the window right now because
I'm cold''), a process known as \textit{pragmatic reasoning}. Theories of pragmatics frame
the process of language comprehension as inference about the
generating goal of an utterance given a rational speaker
\cite{grice1975,dale1995,frank2012}. For example, a listener might
reason, ``if she had wanted me to think `all' of the cookies, she
would have said `all'---but she didn't. Hence `all' must not be true
and she must have eaten some {\it but not all} of the cookies.'' This kind of reasoning is core to language use.

But pragmatic reasoning about meaning-in-context relies on stable literal meanings that must  themselves be learned.
%and often intersects with learning the meaning of language in general. 
In both adults and children, uncertainty about word meanings is
common, and often considering speakers' pragmatic goals can help to resolve this uncertainty. For example,
if a novel word is used in a context containing both a novel and a
familiar object, young children can make the inference
that the novel word refers to the novel object \cite{markman1988}.\footnote{Very young children
  make inferences that are often labeled as ``pragmatic'' in that they
  involve reasoning about context \cite{clark1988,baldwin1993}, though
  they also show weaknesses in specific inferences (e.g.~strengthening
  {\sc some} to {\sc some but not all} \cite{papafragou2003}). Here we
  remain agnostic about the age at which children are able to make
  such inferences robustly, as it may vary depending on the linguistic
  materials being used in the inference \cite{barner2011}.} For adults
who are proficient language users, there are also a variety of intriguing cases
in which listeners seem to create situation- and task-specific ways of
referring to particular objects. For example, when asked to refer to
idiosyncratic geometric shapes, over the course of an experimental
session, participants create conventionalized descriptions that allow
them to perform accurately even though they do not begin with a shared
label \cite{krauss1964,clark1986}. In both of these examples, reasoning about another person's goals informs language learners' estimates of what words are likely to mean.

@@

Formally, let an unadorned $L$ and $S$ denote our convention-based
listener and speaker. By assumption (c), the top-level listener has uncertainty over the lexicon; if the lexicon were known they would draw inferences as in $L_n$ above:
%\begin{align*}
%  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
%    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior. \\
%    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
%  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
%\end{align*}
%
\begin{align*}
  P_L(\obj | \word, \text{$L$'s data}) &=  \sum_\lex P_{L_n}(\obj | \word, \lex) P(\lex | \text{$L$'s data}) \\
\end{align*}
Here $L$'s data consists of $S$'s utterances together with whatever
other information $L$ may have about their meaning (e.g.~from
contextual clues). 
By assumption (b), $L$ treats these  utterances
as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus
as being informative about the lexicon. For instance, when the data is a set of fully observed 
word-referent pairs $\{w_i, o_i\}$:
\begin{align*}
  P(\lex | \text{$L$'s data}) \propto P(\lex) \prod_i P_{S_{n-1}}(w_i|o_i,  \lex)
\end{align*}
The top-level speaker attempts to select the word which soft-maximizes their utility, with utility now being defined in terms of the informativity of the expectation (over lexicons) that the listener will have for at the right referent\footnote{An alternative model would have the speaker take the expectation over informativity, instead of the informativity of the expectation.}:
\begin{align*}
  P_S(\word | \obj, \text{$S$'s data}) &\propto \exp\Big(\lambda \big(\log  \sum_\lex P_{L_{n-1}}(\obj | \word, \lex) P(\lex|\text{$S$'s data}) - \text{cost}(\word)\big)\Big)
\end{align*}
Here $P(\lex|\text{$S$'s data})$ is defined similarly, when $S$ observes
$L$'s interpretations of various utterances, and treats them as
samples from $L_n$, not $L$.

@@

Many sets of words in natural language form scales in which each term
is more informative than, and is entailed by, the previous. ``Some'' and ``all'' form a scale of this
type. While ``I ate some of the cookies'' is compatible with the
followup ``in fact, I ate {\it all} of the cookies,'' the reverse is
not true. ``Might'' and ``must'' are another example, as are ``OK,''
``good,'' and ``excellent.'' All of these scales allow for {\it scalar
  implicatures} \cite{grice1975}: if a speaker uses a weaker term, a
listener is licensed to infer that the stronger term was not true in
this situation. So although ``I ate some of the cookies'' could in
principle be compatible with eating {\sc all} of them, the listener is lead
to believe that {\sc some-but-not-all}' is the likely state of
affairs. We use the term {\it specificity implicature} here (following
\cite{bergen2012}) to denote that we are agnostic about whether such
implicatures are conventionalized or whether they use a
contextually-defined scale.

The recursive pragmatic reasoning portions of our model capture
findings on specificity implicature. Consider a representation of
scalar implicature as a matrix $\left[
    \begin{array}{cc}
      0.5 & 0.5 \\
      0 & 1
    \end{array}
  \right]$, where rows show the words ``some'' and ``all'' and columns
  indicate worlds where {\sc some but not all} and {\sc all} are true,
  respectively, capturing the regularity that ``some'' can be said of
  situations where {\sc all} is actually the case. Assuming uniform
  costs for the two words, a rational speaker confronted with such a
  lexicon will prefer to say ``some'' only when they mean \textsc{some
    but not all}, and this is therefore the interpretation that a
  rational listener will place on ``some'' as well. To simulate this
  situation, we place a Dirichlet(10, 10) prior on the meaning of
``some'' and a Dirichlet(1, 10) prior on the meaning of
``some'', giving speaker probabilities of
  $\left[
    \begin{array}{cc}
      1.00 & 0.00 \\
      0.00 & 1.00
    \end{array}
  \right]$
and listener probabilities of
$\left[
    \begin{array}{cc}
      .86 & .14 \\
      .03 & .97
    \end{array}
  \right]$. These revised interpretation probabilities minimize
  ambiguity (with weight concentrated on the diagonal), capturing the specificity implicature.

@@

If each agent in a multi-agent setting assumes the existence of a
shared, conventional norm underlying everyone's behaviour, and further
assumes that other agents are knowledgeable about this norm while they
themselves are not, then we are able to simultaneously preserve most
of the normative Bayesian framework, explicitly model learning and
recursive social reasoning within the same framework, and avoid
paradox. The resulting model unifies two previously separate strands
of modelling work on language usage and acquisition, and -- having
done so -- accounts for not just the phenomena these previous models
accounted for, but a variety of new ones as well. We propose new
explanations of the child language acquisition phenomena of
disambiguation without mapping \cite{horst2008} and the acquisition of
scalar adjectives (XX ask Mike what the levinson cite is really, see
also table), and demonstrate that our model is capable of developing
novel and efficient communicative systems through iterated learning
within the context of a single simulated conversation.

Accomplishing this requires limited deviations from rationality, and
as such necessarily introduces biases in agent behaviour. We've
identified two such biases here: the tendency for pragmatic speakers
and listeners to accentuate useful, sparse patterns in their
communicative systems (\S\ref{sec:emergence}), and for cheap words to
be assigned to common objects (\S\ref{sec:horn-emergence}).
Remarkably, while both are locally sub-optimal for individual agents,
they both systematically drive the overall communicative system
towards greater global efficiency. In the long term, these processes
should leave their mark on the structure of the language itself, which
may provide a partial explanation for how languages become so tightly
optimized for effective communication \cite{zipf1936,piantadosi2011}.

>>>>

Natural language is a powerful tool for conveying meaning in
context. Consider the utterances
``it's raining,'' ``I ate some of the cookies,'' or ``can you close
the window?'' In each, a listener must go beyond the literal meaning
of the words to fill in contextual details (``it's raining here and
now''), infer that a stronger alternative is not true (``I ate some
but not all of the cookies''), or more generally infer the speaker's
communicative goal (``I want you to close the window right now because
I'm cold''), a process known as \textit{pragmatic reasoning}. Theories of pragmatics frame
the process of language comprehension as inference about the
generating goal of an utterance given a rational speaker
\cite{grice1975,dale1995,frank2012}. For example, a listener might
reason, ``if she had wanted me to think `all' of the cookies, she
would have said `all'---but she didn't. Hence `all' must not be true
and she must have eaten some {\it but not all} of the cookies.'' This kind of reasoning is core to language use.

But pragmatic reasoning often intersects with language learning as
well. In both adults and children, uncertainty about word meanings is
common, and often considering speakers' pragmatic goals can help to resolve this uncertainty. For example,
if a novel word is used in a context containing both a novel and a
familiar object, young children can make the inference
that the novel word refers to the novel object \cite{markman1988}.\footnote{Very young children
  make inferences that are often labeled as ``pragmatic'' in that they
  involve reasoning about context \cite{clark1988,baldwin1993}, though
  they also show weaknesses in specific inferences (e.g.~strengthening
  {\sc some} to {\sc some but not all} \cite{papafragou2003}). Here we
  remain agnostic about the age at which children are able to make
  such inferences robustly, as it may vary depending on the linguistic
  materials being used in the inference \cite{barner2011}.} For adults
who are proficient language users, there are also a variety of intriguing cases
in which listeners seem to create situation- and task-specific ways of
referring to particular objects. For example, when asked to refer to
idiosyncratic geometric shapes, over the course of an experimental
session, participants create conventionalized descriptions that allow
them to perform accurately even though they do not begin with a shared
label \cite{krauss1964,clark1986}. In both of these examples, reasoning about another person's goals informs language learners' estimates of what words are likely to mean.

@@

Formally, let an unadorned $L$ and $S$ denote our convention-based
listener and speaker. By assumption (c), the top-level listener has uncertainty over the lexicon; if the lexicon were known they would draw inferences as in $L_n$ above:
%\begin{align*}
%  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
%    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior. \\
%    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
%  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
%\end{align*}
%
\begin{align*}
  P_L(\obj | \word, \text{$L$'s data}) &=  \sum_\lex P_{L_n}(\obj | \word, \lex) P(\lex | \text{$L$'s data}) \\
\end{align*}
Here $L$'s data consists of $S$'s utterances together with whatever
other information $L$ may have about their meaning (e.g.~from
contextual clues). 
By assumption (b), $L$ treats these  utterances
as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus
as being informative about the lexicon. For instance, when the data is a set of fully observed 
word-referent pairs $\{w_i, o_i\}$:
\begin{align*}
  P(\lex | \text{$L$'s data}) \propto P(\lex) \prod_i P_{S_{n-1}}(w_i|o_i,  \lex)
\end{align*}
The top-level speaker attempts to select the word which soft-maximizes their utility, with utility now being defined in terms of the informativity of the expectation (over lexicons) that the listener will have for at the right referent\footnote{An alternative model would have the speaker take the expectation over informativity, instead of the informativity of the expectation.}:
\begin{align*}
  P_S(\word | \obj, \text{$S$'s data}) &\propto \exp\Big(\lambda \big(\log  \sum_\lex P_{L_{n-1}}(\obj | \word, \lex) P(\lex|\text{$S$'s data}) - \text{cost}(\word)\big)\Big)
\end{align*}
Here $P(\lex|\text{$S$'s data}$ is defined similarly, when $S$ observes
$L$'s interpretations of various utterances, and treats them as
samples from $L_n$, not $L$.

@@

Many sets of words in natural language form scales in which each term
refers to successively stronger restrictions on the situations
compatible with the term. ``Some'' and ``all'' form a scale of this
type. While ``I ate some of the cookies'' is compatible with the
followup ``in fact, I ate {\it all} of the cookies,'' the reverse is
not true. ``Might'' and ``must'' are another example, as are ``OK,''
``good,'' and ``excellent.'' All of these scales allow for {\it scalar
  implicatures} \cite{grice1975}: if a speaker uses a weaker term, a
listener is licensed to infer that the stronger term was not true in
this situation. So although ``I ate some of the cookies'' could in
principle be compatible with eating {\sc all} of them, the listener is lead
to believe that {\sc some-but-not-all}' is the likely state of
affairs. We use the term {\it specificity implicature} here (following
\cite{bergen2012}) to denote that we are agnostic about whether such
implicatures are conventionalized or whether they use a
contextually-defined scale.

The recursive pragmatic reasoning portions of our model capture
findings on specificity implicature. Consider a representation of
scalar implicature as a matrix $\left[
    \begin{array}{cc}
      0.5 & 0.5 \\
      0 & 1
    \end{array}
  \right]$, where rows show the words ``some'' and ``all'' and columns
  indicate worlds where {\sc some but not all} and {\sc all} are true,
  respectively, capturing the regularity that ``some'' can be said of
  situations where {\sc all} is actually the case. Assuming uniform
  costs for the two words, a rational speaker confronted with such a
  lexicon will prefer to say ``some'' only when they mean \textsc{some
    but not all}, and this is therefore the interpretation that a
  rational listener will place on ``some'' as well.

% To simulate this
%   situation, we place a Dirichlet(10, 10) prior on the meaning of
% ``some'' and a Dirichlet(1, 10) prior on the meaning of
% ``some'', giving speaker probabilities of
%   $\left[
%     \begin{array}{cc}
%       1.00 & 0.00 \\
%       0.00 & 1.00
%     \end{array}
%   \right]$
% and listener probabilities of
% $\left[
%     \begin{array}{cc}
%       .86 & .14 \\
%       .03 & .97
%     \end{array}
%   \right]$. These revised interpretation probabilities minimize
%   ambiguity (with weight concentrated on the diagonal), capturing the specificity implicature.

@@

Language learners and language users must consider word meanings both
within and across contexts. A critical part of this process is
reasoning pragmatically about agents' goals in individual
situations. In the current work we assume that agents communicating
with one another assume that there is a shared conventional lexicon
which they both rely on. They then
reason recursively about how this lexicon should be used to convey
particular meanings in context. This pair of assumptions allows us to
create a model that unifies two previously separate strands
of modeling work on language usage and acquisition, and account for a
variety of new phenomena. In particular, we consider new
explanations of disambiguation in early word learning and the acquisition of
scalar adjectives, and demonstrate that our model is capable of developing
novel and efficient communicative systems through iterated learning
within the context of a single simulated conversation.

Our work identifies two biases in communicators' behavior: a tendency for pragmatic speakers
and listeners to accentuate useful, sparse patterns in their
communicative systems (\S\ref{sec:emergence}), and for short,
high-frequency (``low cost'') expressions to
be assigned to common objects (\S\ref{sec:horn-emergence}).
Remarkably, while both of these biases are locally sub-optimal for individual agents,
they both systematically drive the overall communicative system
towards greater global efficiency. In the long term, these processes
should leave their mark on the structure of the language itself, which
may contribute to explaining how languages become 
optimized for effective communication \cite{zipf1936,piantadosi2011}.
